Stochastic Approximation Monte Carlo with a Dynamic Update Factor (SAD)

### 2019-02-11 papers/histogram/sad.tex
minor edits to sad paper

\begin{figure}
  \includegraphics[width=\columnwidth]{figs/N50-lndos-comparison}
  \caption{The entropy of a square well fluid with 50 atoms and filling fraction
        0.3 as the green line.  The green hatched area reflects the
        minimum entropy change needed to converge to the true value.
        The light blue area is the quadratic approximation
        for the change in entropy.  The vertical dotted lines represent
        the energy corresponding to $T=0.\overline{3}$ and $T=\infty$.}
\end{figure}

While for SAMC the update factor is defined in the original
implementation, for SAD the update factor $\gamma_{t}^{\text{SAD}}$ is
thought of as $\nicefrac{\text{d}S}{\text{d}t}$. This tells us that
the SAMC parameter
$t_0$ should have dimensions of entropy.
We begin with an estimate of the average value of the entropy (relative
to the lowest entropy at $T_{\min}$).  If we assume a quadratic
dependence on energy, this is given by
\begin{align}
\langle S\rangle \approx \frac13 \frac{E({T=\infty}) - E(T_{\min})}{T_{\min}}
\end{align}
We approximate this energy difference by
$E_H -E_L$ where $E_H$ and $E_L$ are defined below.
The entropy numerator of the update factor
in general should scale with the total number of interesting
energy states $N_S$, since updates to the weights are distributed between
that many energy states.  The product $N_S\langle S\rangle$ is the
total change of entropy required (starting from constant weights) to find
the true entropy, and puts a lower bound on the convergence time.
After \emph{long} times, when
all the energies have been long ago found,
we wish for a lower $t_0$ in order to more rapidly refine the remaining
error in entropy.  We enable this
by tracking the last time we encountered a new energy $t_L$, and
gradually transition to lower $t_0$ (but still asymptotically scaling
as $\gamma_t \propto 1/t$ to ensure eventual convergence).  Finally,
we wish for an update factor that is \emph{never} greater than 1, because
a very large update factor could introduce errors in entropy that may take
many iterations to remove.  The SAD
expression for $\gamma_t$ which incorporates these ideas is:
\begin{align}
  \gamma_{t}^{\text{SAD}} =
     \frac{
       \frac{E_{H}-E_{L}}{T_{\text{min}}} + \frac{t}{t_L}
     }{
       \frac{E_{H}-E_{L}}{T_{\text{min}}} + \frac{t}{N_S}\frac{t}{t_L}
     }
\end{align}
This factor asymptotically has the same $1/t$ behavior as the original
SAMC algorithm;
however now every time a new energy is found to be important
during the course of
the simulation, the update factor will experience a jump. This behavior
allows SAD to dynamically prevent the update factor from decreasing too
rapidly.

@@

Figures~\ref{fig:N50-ff0.3-avg-error}
and~\ref{fig:N50-ff0.3-avg-error-slow} show the average entropy vs
moves for a system consisting of 50 atoms that are placed in a periodic
simple cubic unit cell with lattice constant chosen to give a filling
fraction of $\eta = 0.3$.  These particles are randomly inserted into
the system and random moves are proposed (with a translation scale
$\delta_0 = 0.05\sigma$ and $\delta_0 = 0.005\sigma$ respectively). For
the $\eta = 30\%$ and $N = 50$ system, we explore the choice of
$\delta_0$ as a parameter.

{\color{red}
$S_{\max} = 120$ and $S_{\min} = 248$ for this system.  The number of
important energy states for this system is then $N_S = S_{\min} - S_{\max} = 145$.
The reduced temperature of $1/3$ is chosen in order to examine the convergence
properties of SAD in a reasonable simulation time.

>>>>

\begin{figure}
  \includegraphics[width=\columnwidth]{figs/N50-lndos-comparison}
  \caption{The entropy of a square well fluid with 50 atoms and filling fraction
        0.3 as the green line.  The green hatched area reflects the
        minimum entropy change needed to converge to the true value.
        The light blue area is the quadratic approximation
        for the change in entropy.  The vertical dotted lines represent
        the energy corresponding to $T=1/3$ and $T=\infty$.}
  \label{fig:entropy-cartoon}
\end{figure}

While for SAMC the update factor is defined in the original
implementation, for SAD the update factor $\gamma_{t}^{\text{SAD}}$ is
thought of as $\nicefrac{\text{d}S}{\text{d}t}$. This tells us that
the SAMC parameter
$t_0$ should have dimensions of entropy.
We begin with an estimate of the average value of the entropy (relative
to the lowest entropy at $T_{\min}$).  If we assume a quadratic
dependence on energy (see Fig.~\ref{fig:entropy-cartoon}), this is given by
\begin{align}
\langle S\rangle \approx \frac13 \frac{E({T=\infty}) - E(T_{\min})}{T_{\min}}
\end{align}
We approximate this energy difference by
$E_H -E_L$ where $E_H$ and $E_L$ are defined below.
The entropy numerator of the update factor
in general should scale with the total number of interesting
energy states $N_S$, since updates to the weights are distributed between
that many energy states.  The product $N_S\langle S\rangle$ is the
total change of entropy required (starting from constant weights) to find
the true entropy, and puts a lower bound on the convergence time.
After \emph{long} times, when
all the energies have been long ago found,
we wish for a lower $t_0$ in order to more rapidly refine the remaining
error in entropy.  We enable this
by tracking the last time we encountered a new energy $t_L$, and
gradually transition to lower $t_0$ (but still asymptotically scaling
as $\gamma_t \propto 1/t$ to ensure eventual convergence).  Finally,
we wish for an update factor that is \emph{never} greater than 1, because
a very large update factor could introduce errors in entropy that may take
many iterations to remove.  The SAD
expression for $\gamma_t$ which incorporates these ideas is:
\begin{align}
  \gamma_{t}^{\text{SAD}} =
     \frac{
       \frac{E_{H}-E_{L}}{T_{\text{min}}} + \frac{t}{t_L}
     }{
       \frac{E_{H}-E_{L}}{T_{\text{min}}} + \frac{t}{N_S}\frac{t}{t_L}
     }
\end{align}
This factor asymptotically has the same $1/t$ behavior as the original
SAMC algorithm;
however now every time a new energy is found to be important
during the course of
the simulation, the update factor will experience a jump. This behavior
allows SAD to dynamically prevent the update factor from decreasing too
rapidly.

@@

Our smaller system consists of 50 square-well atoms with a well width
of $\lambda=1.3$, in a volume corresponding to a filling fraction of
$\eta=0.3$.  For this simulation, we chose a minimum reduced
temperature of $1/3$, which corresponds to an interesting energy range
from -248 to -120.  The entropy of this system is shown in
Fig.~\ref{fig:entropy-cartoon} above, which shows that over this
energy range the entropy differs by 198.

Figures~\ref{fig:N50-ff0.3-avg-error}
and~\ref{fig:N50-ff0.3-avg-error-slow} show the average error in the entropy vs
moves for a system consisting of 50 atoms that are placed in a periodic
simple cubic unit cell with lattice constant chosen to give a filling
fraction of $\eta = 0.3$.  These particles are randomly inserted into
the system and random moves are proposed (with a translation scale
$\delta_0 = 0.05\sigma$ and $\delta_0 = 0.005\sigma$ respectively). For
the $\eta = 30\%$ and $N = 50$ system, we explore the choice of
$\delta_0$ as a parameter.

The interesting range of energies for this system is from
{\color{blue}
$S_{\max} = 120$ and $S_{\min} = 248$ for this system.  The number of
important energy states for this system is then $N_S = S_{\min} - S_{\max} = 145$.
The reduced temperature of $1/3$ is chosen in order to examine the convergence
properties of SAD in a reasonable simulation time.
